On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-16 paper/application.tex

\section{Application to protoplanetary disks}\label{2dppd}
We consider the stability of realistic, self-gravitating
protoplanetary disks. Specifically, we will apply our linear framework
to the gravito-turbulent disk model recently developed by 
\cite{rafikov15}. This 2D, Keplerian disk orbits a Solar mass star and 
is defined by the following parameters,

\begin{itemize}
  \item $\dot{M}$, the global radial accretion rate;
  \item $Q_0$, the value of the 2D Toomre parameter where the disk is
    gravito-turbulent;
  \item $T_\mathrm{irr}$, the irradiation or floor temperature;
  \item $\alpha_m$, the dimensionless viscosity associated with other
    sources of turbulence.
\end{itemize} 
These are used in the statements of thermal equilibrium, mass and
angular momentum conservation, together with an opacity law, 
to construct a global disk model with surface density $\Sigma(R)$ and
temperature $T(R)$ where $R$ is the global cylindrical radius from the
star. This gives the $Q(R)$ value required for input
into our linear framework. We derive other dimensionless
parameters below.

@@

\subsection{Inviscid stability condition} 
%We specialize to $a=0, b=2$ for the opacity law. 
We note that for the adopted opacity law, Eq. \ref{opacity_law}, 
\begin{align*}
  \theta = \frac{4-g}{4-2g},
\end{align*}
implying $5/6<\theta<3/2$. 
%For a non-irradiated disk, $T_\mathrm{irr}=0$, $\theta\to5/6$ as
%$\tau\to0$ and $\theta\to3/2$ as $\tau\to\infty$.  
From our discussion in \S\ref{2d_inviscid}, we conclude that this
system cannot be unstable without
viscous effects acting on the perturbations if 
\begin{align}
  \gamma > \frac{3}{2} \quad \text{and} \quad Q > \sqrt{\frac{6}{5}} \label{ppd_invisc_cond}
\end{align} 
are both satisfied.

\subsection{Example 2D calculation}\label{pp2d_example}
We consider a disk with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. For the adopted opacity
law we follow \citeauthor{rafikov15} and take $\kappa_{d0} =
5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$. Here we use
$\gamma=1.6$ so the inviscid stability condition
(Eq. \ref{ppd_invisc_cond}) is satisfied. 
%{\bf actually this opacity law only valid $>30$au for this disk}

Fig. \ref{rafikov_model} shows the equilirbium disk profile in terms
of $Q$, $\alpha$, $\beta$, and $\theta$. These are set as input to the
2D dispersion relation to yield a growth timescale, in units of the
local orbital period, shown in Fig. \ref{rafikov_growth}. We also plot
an analytic estimate based on Eq. \ref{gammie_smallk} (instead of
Eq. \ref{gammie_maxrate_simple} since here $\theta\sim 1$) which gives 
\begin{align}
  S = \frac{27\alpha}{16\theta^3Q^4}, 
\end{align}
and the optimum wavenumber is
\begin{align}
  |K| = \frac{3}{2\theta Q}.
\end{align}
%For the
%stability calculation we use $\mu=-1, \lambda=0$ for the viscosity
%prescription.

The disk becomes dynamically unstable with growth timescales $\lesssim
10$ orbits for $R\gtrsim60$AU. Coincidentally, beyond this radius
$\alpha\gtrsim 0.1$ and $\beta\lesssim 3$, which is often quoted as 
criteria for disk fragmentation \citep[e.g.][]{rafikov15}. 
%{\bf despite apply beta cooling results to realistic cooling disks}
Our result suggest an alternative interpretation of disk fragmentation
is where the gravito-turbulent state becomes dynamically unstable due to
both viscosity and self-gravity.

Fig. \ref{rafikov_model} shows that the most unstable wavelength is a
few times the disk thickness. The increase in $|K|$ from $\sim
10\mathrm{AU}$ to $\sim 20\mathrm{AU}$ is due to the decrease in $Q$;
while that from $\sim 60\mathrm{AU}$ to $\sim 100\mathrm{AU}$ occurs
as the disk transitions from optically thick to optically thin
regimes. The mismatch at large distances is not surprising since the
above expressions assume $|K|\ll 1$. Nevertheless the analytic
estimates reproduce the correct behavior.

@@

\subsection{3D PPD}
%{\bf radiative diffusion only for optically thick disk. how to connect
%to optically thin regions?}
We briefly consider 3D PPDs subject to explicit radiative diffusion 
(\S\ref{rad_cool}). Given the $\alpha(R)$ and $Q(R)$ profiles obtained
from the 2D model above, at each radius $R$ we obtain the vertical
structure from Eq. \ref{vert_eq1}---\ref{thermal_eq}, with 
Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux.
We then solve the 3D eigenvalue problem as in \S\ref{3ddisk}, with the
additional boundary condition $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
We use the same disk model as in \S\ref{pp2d_example} but without 
irradiation ($T_\mathrm{irr}=0$), and a slightly smaller vertical
domain with $\rho(\zmax)=0.1\rho_0$.  %to avoid very low density

>>>>

\section{Application to protoplanetary disks}\label{2dppd}
We now consider the stability of realistic, self-gravitating
protoplanetary disks. Specifically, we will apply our linear framework
to the gravito-turbulent disk model recently developed by 
\cite{rafikov15}. This 2D, Keplerian disk orbits a Solar mass star and 
is defined by the following parameters,

\begin{itemize}
  \item $\dot{M}$, the global radial mass accretion rate;
  \item $Q_0$, the value of the 2D Toomre parameter where the disk is
    gravito-turbulent;
  \item $T_\mathrm{irr}$, the irradiation or floor temperature;
  \item $\alpha_m$, the dimensionless viscosity associated with other
    sources of turbulence.
\end{itemize} 
These are used in the statements of thermal equilibrium, mass and
angular momentum conservation, together with an opacity law, 
to construct a global disk model with surface density $\Sigma(R)$ and
temperature $T(R)$ where $R$ is the global cylindrical radius from the
star. \citep[See][for details.]{rafikov15} This gives the $Q(R)$
value required for input into our linear framework. We derive other
dimensionless parameters below.

@@

\subsection{Inviscid stability condition} 
%We specialize to $a=0, b=2$ for the opacity law. 
We note that for the adopted opacity law, Eq. \ref{opacity_law}, 
\begin{align*}
  \theta = \frac{4-g}{4-2g},
\end{align*}
implying $5/6<\theta<3/2$. 
%For a non-irradiated disk, $T_\mathrm{irr}=0$, $\theta\to5/6$ as
%$\tau\to0$ and $\theta\to3/2$ as $\tau\to\infty$.  
From the results in \S\ref{2d_inviscid}, we conclude that this
system cannot be unstable anywhere without
viscous effects acting on the perturbations if 
\begin{align}
  \gamma > \frac{3}{2} \quad \text{and} \quad Q > \sqrt{\frac{6}{5}} \label{ppd_invisc_cond}
\end{align} 
are both satisfied.

A PPD becomes optically-thin ($\tau\to 0$) at
large radii, and $\theta$ decreases. The first inequality in
Eq. \ref{ppd_invisc_cond} then relaxes to $\gamma > 5/6$ in the
outer disk, which is always true. On the other hand, gravito-turbulent
simulations typically show $Q\simeq 1.5$---2, which satisfies the
second critrion. This means that in the
outer regions of a realistic PPD, cooling itself cannot cause instability.

%conditions are likely satisfied in the outer disk 
%Condition \ref{ppd_invisc_cond} must be violated in order for cooling
%to destablize

\subsection{Example 2D calculation}\label{pp2d_example}
We consider a disk with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. For the adopted opacity
law we follow \citeauthor{rafikov15} and take $\kappa_{d0} =
5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$. Here we use
$\gamma=1.6$ so the inviscid stability condition
(Eq. \ref{ppd_invisc_cond}) is satisfied. Then cooling alone cannot
destabilize the disk. Fig. \ref{rafikov_model} shows the equilirbium
disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
These are set as input to the 2D dispersion relation.

Fig. \ref{rafikov_growth} show growth timescales and
optimum wavenumbers for viscous GI in this disk model.  
We also plot analytic estimates based on Eq. \ref{gammie_smallk} (instead of
Eq. \ref{gammie_maxrate_simple} since here $\theta\sim 1$) 
which gives the optimum wavenumber as
\begin{align}
  |K| = \frac{3}{2\theta Q},
\end{align}
with growth rate
\begin{align} 
  S = \frac{27\alpha}{16\theta^3Q^4}. 
\end{align}
These are similar to the isothermal results of
\citet[][their Eq. 19 and 21, respectively]{sterzik95}, and identical if one
takes $\theta=1$.

The most unstable wavelength is a few times the disk thickness. The
increase in $|K|$ from $\sim 10\mathrm{AU}$ to $\sim 20\mathrm{AU}$ is
due to the decrease in $Q$; while that from $\sim 60\mathrm{AU}$ to
$\sim 100\mathrm{AU}$ occurs as the disk transitions from optically
thick to optically thin regimes. The mismatch at large distances is
not surprising since the above expressions assume $|K|\ll
1$. Nevertheless the analytic estimates reproduce the correct
behavior.

The disk is subject to viscous GI on dynamical timescales ($\lesssim
10$ orbits) for $R\gtrsim60$AU. Coincidentally, beyond this radius
$\alpha\gtrsim 0.1$ and $\beta\lesssim 3$, which is often quoted as 
criteria for disk fragmentation \citep[e.g.][]{rafikov15}. 
This suggests that fragmentation in the outer parts of a realistic PPD
occurs is physically due to large turbulent stresses.

%{\bf despite apply beta cooling results to realistic cooling disks}
%Our result suggest an alternative interpretation of disk fragmentation
%is where the gravito-turbulent state becomes dynamically unstable due to
%both viscosity and self-gravity.

@@

\subsection{3D PPD with radiative diffusion}
%{\bf radiative diffusion only for optically thick disk. how to connect
%to optically thin regions?}
We briefly consider 3D PPDs subject to explicit radiative diffusion 
(\S\ref{rad_cool}). Given the $\alpha(R)$ and $Q(R)$ profiles obtained
from the 2D model above, at each radius $R$ we obtain the vertical
structure from Eq. \ref{vert_eq1}---\ref{thermal_eq}, with 
Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux.
We then solve the 3D eigenvalue problem as in \S\ref{3ddisk}, with the
additional boundary condition $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
We use the same disk model as in \S\ref{pp2d_example} but without 
irradiation ($T_\mathrm{irr}=0$), and a slightly smaller vertical
domain with $\rho(\zmax)=0.1\rho_0$.  %to avoid very low density

### 2016-01-16 paper/results2d.tex

\section{Two-dimensional disks with beta cooling}\label{result_2d}
We begin in the 2D limit with beta cooling  to make connection 
with previous studies. The disk
material assumed to be confined to the midplane so $\delta v_z=0$. 
We make the replacements 
$\rho \to \Sigma$ and $P\to\Pi$, the vertically-integrated
pressure, and set $\Gamma=1$ without loss of generality. 
%The perturbations $\delta\Sigma$, 
%$\delta\Pi$, $\delta v_x$ and $\delta v_y$ have no vertical
%dependence and $\delta v_z=0$. 
The gravitational potential perturbation remains 3D and its midplane
value is given by    
\begin{align}
  \delta \Phi(z=0) = -\frac{2 \pi G}{|k|}\delta\Sigma
\end{align}
\citep{shu70}.

We obtain from the linearized equations 
an algebraic dispersion relation $s = s(k)$. It is simplest
to write this  
in terms of the dimensionless growth rate $S = s/\Omega$ and
wavenumber $K=kH = k c_{s0}/\Omega$ as
\begin{align}\label{thindisk}
  f(S,K)\equiv AD - BC = 0,   
\end{align}
where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp}.

\subsection{Inviscid limit}\label{2d_inviscid}
Here we neglect viscosity by setting 
$\alpha = \alpha_b = 0$. This implies an external,
time-independent heat source must be invoked to balance the imposed
cooling rate in equilibrium ($\mathcal{H}_\mathrm{ext}\neq 0$). Alternatively, we are assuming 
viscosity is only important for setting up thermal equilibrium, and
can be neglected for the pertubations. This eliminates viscous GI and 
allows us to quantify the \emph{sole} effect of cooling.

Eq. \ref{thindisk} simplifies to  
\begin{align}\label{inviscid}
%  &\beta S^3 + S^2 - \beta \left\{ \frac{2|K|}{\qtwo} - \left[2(2-q) +
%    \gamma K^2\right]\right\}S \notag\\ &- \left[\frac{2|K|}{\qtwo} - 2(2-q)\right]
%  = 0. 
  S^2 = \frac{2|K|}{Q} - 2(2-q) - \left(\frac{\theta + \beta \gamma
    S}{1+\beta S}\right)K^2, 
\end{align}
similar to the classic Lin-Shu dispersion relation. The first term on
the right-hand-side represents destabilization by self-gravity; 
the second and third terms represent stabilization by rotation and
pressure, respectively. The imposed cooling term only affects the
pressure response.

Eq. \ref{inviscid} is a cubic equation in $S$. Applying the
Routh-Hurwitz criterion, we find that stability is ensured if 
\begin{align}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. Thus, the question of stability does not depend on
the cooling rate $\beta$ (provided it is non-zero and finite). The
system becomes more unstable with decreasing $\theta$.

Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p |K| =
0$ and $S = 
S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
\begin{align}\label{kstar}
%  |K_*| = \frac{1}{\gamma \qtwo} \left(1 + \frac{1}{\beta S_*}\right). 
  |K_*| = \frac{1+\beta S_*}{Q\left(\theta + \gamma \beta S_*\right)}
\end{align}
Inserting this into Eq. \ref{inviscid}, we find the maximum growth
rate satisfies
\begin{align}\label{inviscid_max}
%  S_*^3 + \left[2(2-q)  - \frac{1}{\gamma \qtwo^2}\right]S_* - \frac{1}{\gamma
%    \qtwo^2 \beta} = 0.
  S_*^2 = \frac{1+\beta S_*}{Q^2\left(\theta + \beta\gamma S_*\right)}
  - 2(2-q).
\end{align}
General solutions to this cubic for $S_*$ are unwieldy, but simple
results may be obtained in special cases, discussed next. 
%We plot examples of the maximimum growth rate as a function of cooling
%times in Fig. \ref{2d_inviscid}. General expressions for the solution
%of Eq. \ref{inviscid_max} are unwieldy, but simple results may 
%be obtained for special cases, as discussed below.  
%\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0cm 0cm]{figures/inviscsg}
%  \caption{Maximum growth rate $s$ in the inviscid limit,
%    obtained from Eq. \ref{inviscid}, as a function of the cooling
%    time $\beta$ for $Q/Q_\mathrm{crit}=1,\,2,\,10$, where $\qcrit$ is
%    defined in Eq. \ref{qcrit_def} as 
%    the value of $Q$ such that the disk is marginally stable in the
%    adiabatic limit. 
%    \label{2d_inviscid}}
%\end{figure}

\subsubsection{$\theta = 0$ and $Q=Q_\mathrm{crit}$}  
Beta cooling with $\theta=0$ is the standard choice for numerical
simulations of self-gravitating disks \citep{gammie01}. There, 
a steady state corresponds to the imposed cooling entirely balanced by heat generated through 
gravito-turbulence. In the present framework, this gravito-turbulence 
is assumed to provide a constant 
background heating ($\mathcal{H}_\mathrm{ext}$) and does
not enter the linearized equations. Cooling, however, is still imposed in the perturbed state. 
For $\theta=0$ the system is unconditionally unstable for
finite $Q$. The most unstable wavelength decreases with decreasing
$\beta$. The $\beta\to0$ limit corresponds to a pressureless disk (not
merely isothermal).

% it is always
%possible to find a sufficiently small scale such that pressure effects
%are

Specifically, let us consider a disk with 
\begin{align}\label{qcrit_def}
  \qtwo = \frac{1}{\sqrt{2\gamma(2-q)}}\equiv Q_{\mathrm{crit}},
\end{align} 
so that it is marginally stable in the absence of cooling.  
How does finite cooling destabilize the disk?  
Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} and setting
$\theta=0$, we find 
%In this case the coefficient of 
%$S_*$ in Eq. \ref{inviscid_max} vanish, so that  
\begin{align}\label{sstar}
  S_*^3 = \left(\frac{1}{\gamma Q_\mathrm{crit}^2 \beta}\right). 
\end{align}
The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
increases with decreasing $\beta$. Instability does not require the
cooling time to be below some critical value.

Nevertheless, we can define a characteristic cooling
time $\beta_*$ as that which removes pressure support against
self-gravity over the only natural lengthscale in the problem, $H$. We
thus set $|K_*|=1$ and find, for the Keplerian disk,

\begin{align}\label{betastar}
  \beta_* = \frac{1}{\left(\sqrt{\gamma} - 1\right)^{3/2}}. 
\end{align}
%Table \ref{bstar_compare} compares Eq. \ref{betastar} to 
%previously reported values of $\beta_c$ 
%for which fragmentation
%occurs.
It is interesting to note that this $\beta_*$ is similar to previously reported 
values of the cooling times, $\beta_c$, below which fragmentation
occurs on a dynamical timescale. We compare these in Table
\ref{bstar_compare}, showing rough agreement.

@@

\subsubsection{Passive disk with $\theta = 1$}
Beta cooling with $\theta=1$ corresponds to 
`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case 
$\beta\to 0$ corresponds to an isothermal disk and $\beta
\to \infty$ corresponds to an adiabatic disk. (These two limits are
equivalent for $\gamma=1$.) The most unstable wavenumber and growth
rates vary weakly with $\beta$.

@@

As a numerical example, we consider a model with $\alpha_b=0$ and
$\alpha=\alpha(\beta)$ given by thermal equilibrium,
Eq. \ref{alpha_beta_relation}. Furthermore, we set
\begin{align}
  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
\end{align}
to mimic a basic state maintained by gravito-turbulence where one can 
expect the dimensionless stress $\alpha \sim Q^{-2}$ \citep{lin87}.

Fig. \ref{gammie_rate_plot} show growth rates as a function of the
wavenumber $k$ obtained from the dispersion relation
Eq. \ref{thindisk}. The limiting behavior for small/large $K$ are
well-captured by Eq. \ref{gammie_smallk} and
\ref{gammie_bigk}. Comparing the two panels shows that increasing the
irradiation level ($\theta$) surpress  
small-scale perturbations.

@@

Fig. \ref{gammie_maxrate_plot} show the maximum growth rate as a
function of the cooling time $\beta$ for the case $\theta=0$. 
%We set $\alpha_b=\theta=0$ and 
%consider three cases (top to bottom panels): 
%\begin{enumerate}[(a)]
%  \item $\alpha$, $\beta$ and $Q$ are independent; \label{casea}
%  \item $\alpha=\alpha(\beta)$ from thermal equlibrium
%    (Eq. \ref{alpha_beta_relation}) and $Q$ specified separately; \label{caseb}
%  \item $\alpha(\beta)$ from Eq. \ref{alpha_beta_relation} and
%    assuming $Q = Q_\mathrm{crit}/\sqrt{\alpha}$. This mimics a
%    basic state maintained by 
%    gravito-turbulence where the viscosity coefficient $\alpha \sim
%    Q^{-2}$ \citep{lin87}.
%    \label{gvisc}
%\end{enumerate} 
There is good agreement between numerical results and
Eq. \ref{gammie_maxrate_simple} for $\beta \gtrsim 1$. The assumption
of $|\beta S|\ll 1$ is not well satisfied at smaller $\beta$ which
explains the mismatch there.

Eq. \ref{gammie_maxrate_simple} gives the limiting behavior (for
negligible $\theta$) as 
\begin{align*}
  S_*\propto \begin{cases}
    \frac{\alpha^{1/4}}{Q\beta^{3/4}} \propto \beta^{-3/2} &  \beta
    \gg \alpha, \\
    \frac{1}{Q\sqrt{\alpha}}  \propto \mathrm{const.} & \beta \ll \alpha,
  \end{cases}
\end{align*}
where the second proportionality applies Eq. \ref{alpha_beta_relation}
and \ref{Qalpha}. The disk is formally unstable for all $\beta$, but
the growth timescales are long ($>10$ orbits) for $\beta \gtrsim
20$. This region is marked by the vertical dashed-dotted lines in
Fig. \ref{gammie_maxrate_plot}.

As $\beta\to 0$ the required viscous heating to set up an equilibrium
state diverges ($\alpha\to\infty$). 
In the present model the increased
viscous damping, which is stabilizing, is balanced by the decreasing
(and hence destablizing) $Q$ required to achieve large $\alpha$
(Eq. \ref{Qalpha}).

Physically, however, we expect $\alpha < 1$. In fact, numerical
simulations of self-gravitating disks show there is a maximum $\alpha$
($\sim 0.06$) that can be sustained before fragmentation
\citep{rice11}. We can interpret this result in our linear framework.
%We can use our case (\ref{gvisc}) to interpret this
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$. Fig. \ref{gammie_maxrate_plot} 
shows that such a system is dynamically unstable with 
growth rate $s = O(\Omega)$.  Thus we do not expect rapidly-cooled,
self-gravitating steady disks to exist for extended periods of
time. This is consistent with previous numerical simulations performed
by \cite{lodato05}.

%Case (\ref{gvisc}) is consistent with

\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \caption{Growth rates maximized over wavenumber as a function of cooling time for the
    case shown in Fig. \ref{gammie_rate_plot}. Solid lines are
    computed from the dispersion relation (Eq. \ref{thindisk}), and the dotted line is an
    estimate from Eq. \ref{gammie_maxrate_simple}. The vertical dashed
    line marks the region with $\alpha > 1$, and the vertical
    dashed-dot line marks the region with growth timescales longer
    than dynamical. 
    \label{gammie_maxrate_plot}}
\end{figure}

>>>>

\section{Two-dimensional disks with beta cooling}\label{result_2d}
We begin in the 2D limit with beta cooling  to make connection 
with previous studies. The disk
material is assumed to be confined to the midplane so $\delta v_z=0$. 
We make the replacement  
$\rho \to \Sigma$, re-interpret $P$ as the vertically-integrated
pressure, and set $\Gamma=1$ without loss of generality. 
The gravitational potential perturbation remains 3D and its midplane
value is given by    
\begin{align}
  \delta \Phi(z=0) = -\frac{2 \pi G}{|k|}\delta\Sigma
\end{align}
\citep{shu70}.

The linearized equations yield an algebraic dispersion relation $s =
s(k)$. We write this  
in terms of the dimensionless growth rate $S = s/\Omega$ and
wavenumber $K=kH = k c_{s0}/\Omega$ as
\begin{align}\label{thindisk}
  f(S,K)\equiv AD - BC = 0,   
\end{align}
where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp}.

\subsection{Inviscid limit}\label{2d_inviscid}
Here we neglect viscosity by setting 
$\alpha = \alpha_b = 0$. This eliminates viscous GI and 
allows us to quantify the \emph{sole} effect of cooling on the
perturbations. An external, time-independent heat source should be
invoked to balance the imposed cooling to allow an equilibrium to be
defined ($\mathcal{H}_\mathrm{ext}\neq 0$). Alternatively, we are
assuming viscosity only provides a background heating and does not
play an active role in the perturbed state. %may or may not be true in a real
                                %disk because we are using viscosity
                                %as proxy for turbulence

Eq. \ref{thindisk} simplifies to  
\begin{align}\label{inviscid}
%  &\beta S^3 + S^2 - \beta \left\{ \frac{2|K|}{\qtwo} - \left[2(2-q) +
%    \gamma K^2\right]\right\}S \notag\\ &- \left[\frac{2|K|}{\qtwo} - 2(2-q)\right]
%  = 0. 
  S^2 = \frac{2|K|}{Q} - 2(2-q) - \left(\frac{\theta + \beta \gamma
    S}{1+\beta S}\right)K^2, 
\end{align}
similar to the classic Lin-Shu dispersion relation, which may be
obtained by taking the limit $|\beta S|\to\infty$. 
The first term on
the right-hand-side represents destabilization by self-gravity; 
the second and third terms represent stabilization by rotation and
pressure, respectively. The imposed cooling only affects the
pressure response.

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. Thus, whether or not the system is stable is
independent of the cooling rate $\beta$ (provided it is non-zero and
finite). The system becomes more unstable with decreasing $\theta$.

Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p |K| =
0$ and $S = 
S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
\begin{align}\label{kstar}
%  |K_*| = \frac{1}{\gamma \qtwo} \left(1 + \frac{1}{\beta S_*}\right). 
  |K_*| = \frac{1+\beta S_*}{Q\left(\theta + \gamma \beta S_*\right)}
\end{align}
%Eq. \ref{kstar} describes how the most unstable  mode is affected by
%cooling ($\beta$) and external heating ($\theta$). 
Inserting this into Eq. \ref{inviscid}, we find the maximum growth
rate satisfies
\begin{align}\label{inviscid_max}
%  S_*^3 + \left[2(2-q)  - \frac{1}{\gamma \qtwo^2}\right]S_* - \frac{1}{\gamma
%    \qtwo^2 \beta} = 0.
  S_*^2 = \frac{1+\beta S_*}{Q^2\left(\theta + \beta\gamma S_*\right)}
  - 2(2-q).
\end{align}
General solutions to this cubic for $S_*$ are unwieldy, but simple
results may be obtained in special cases, discussed next.

\subsubsection{$\theta = 0$}  
%Beta cooling with $\theta=0$ is commonly employed in 
%numerical simulations of self-gravitating disks \citep{gammie01}. 
For beta cooling with $\theta=0$, the disk is unconditionally
unstable for finite $Q$. The most unstable wavelength decreases with
decreasing $\beta$. The $\beta\to0$ limit corresponds to a
pressureless disk (not merely isothermal).

Let us consider a disk with 
\begin{align}\label{qcrit_def}
  \qtwo = \frac{1}{\sqrt{2\gamma(2-q)}}\equiv Q_{\mathrm{crit}},
\end{align} 
so that it is marginally stable in the absence of cooling.  
How does finite cooling destabilize the disk?  
Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} and setting
$\theta=0$, we find 
%In this case the coefficient of 
%$S_*$ in Eq. \ref{inviscid_max} vanish, so that  
\begin{align}\label{sstar}
  S_*^3 = \frac{1}{\gamma Q_\mathrm{crit}^2 \beta}. 
\end{align}
The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
increases with decreasing $\beta$. Instability does not require the
cooling time to be below some critical value.

Nevertheless, we can define a characteristic cooling
time $\beta_*$ as that which removes pressure support against
self-gravity over the natural lengthscale in the problem, $H$. We
thus set $|K_*|=1$ and find, for the Keplerian disk,

\begin{align}\label{betastar}
  \beta_* = \frac{1}{\left(\sqrt{\gamma} - 1\right)^{3/2}}. 
\end{align}
%Table \ref{bstar_compare} compares Eq. \ref{betastar} to 
%previously reported values of $\beta_c$ 
%for which fragmentation
%occurs.
This $\beta_*$ is similar to values of the cooling times, $\beta_\mathrm{c}$,
below which fragmentation occur on dynamical timescales in
numerical simulations of self-gravitating disks
\citep{gammie01,rice05,rice11}, which employ the same beta cooling prescription.    
We compare these in Table \ref{bstar_compare}, showing rough
agreement; and further discuss the potential significance of this result in
\S\ref{prev_works}.

@@

\subsubsection{$\theta = 1$}
Beta cooling with $\theta=1$ corresponds to 
`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case 
$\beta\to 0$ corresponds to an isothermal disk and $\beta
\to \infty$ corresponds to an adiabatic disk. (These two limits are
equivalent for $\gamma=1$.) The most unstable wavenumber and growth
rates vary weakly with $\beta$.

@@

As a numerical example, we consider a model with $\alpha_b=0$ and
$\alpha=\alpha(\beta)$ given by thermal equilibrium (
Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
\begin{align}
  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
\end{align}
to mimic a basic state maintained by gravito-turbulence where one
might expect the dimensionless stress $\alpha \sim Q^{-2}$
\citep{lin87}.

Fig. \ref{gammie_rate_plot} show growth rates as a function of the
wavenumber $k$ obtained from the dispersion relation
Eq. \ref{thindisk}. The limiting behavior for small/large $K$ are
well-captured by Eq. \ref{gammie_smallk} and
\ref{gammie_bigk}. Comparing the two panels shows that increasing the
irradiation level ($\theta$) surpresses small-scale perturbations.

@@

Fig. \ref{gammie_maxrate_plot} show the maximum growth rate as a
function of the cooling time $\beta$ for $\theta=0$. 
%We set $\alpha_b=\theta=0$ and 
%consider three cases (top to bottom panels): 
%\begin{enumerate}[(a)]
%  \item $\alpha$, $\beta$ and $Q$ are independent; \label{casea}
%  \item $\alpha=\alpha(\beta)$ from thermal equlibrium
%    (Eq. \ref{alpha_beta_relation}) and $Q$ specified separately; \label{caseb}
%  \item $\alpha(\beta)$ from Eq. \ref{alpha_beta_relation} and
%    assuming $Q = Q_\mathrm{crit}/\sqrt{\alpha}$. This mimics a
%    basic state maintained by 
%    gravito-turbulence where the viscosity coefficient $\alpha \sim
%    Q^{-2}$ \citep{lin87}.
%    \label{gvisc}
%\end{enumerate} 
There is good agreement between numerical results and
Eq. \ref{gammie_maxrate_simple} for $\beta \gtrsim 1$. 
%The assumption
%of $|\beta S|\ll 1$ is not well satisfied at smaller $\beta$ which
%explains the mismatch there. 
Eq. \ref{gammie_maxrate_simple} gives the
limiting behavior for this case as   
\begin{align*}
  S_*\propto \begin{cases}
%    \frac{\alpha^{1/4}}{Q\beta^{3/4}} \propto \beta^{-3/2} 
    \alpha^{1/4}\beta^{-3/4}Q^{-1}\propto \beta^{-3/2} 
    &  \beta \gg \alpha, \\
%    \frac{1}{Q\sqrt{\alpha}}  
    \alpha^{-1/2}Q^{-1} = \mathrm{const.} & \beta \ll \alpha,    
  \end{cases}
\end{align*}
where we have applied Eq. \ref{alpha_beta_relation}
and \ref{Qalpha} to obtain the final result. 
The disk is formally unstable for all $\beta$, but
the growth timescales are long ($>10$ orbits) for $\beta \gtrsim
20$. This region is marked by the vertical dashed-dotted line in
Fig. \ref{gammie_maxrate_plot}.

%As $\beta\to 0$ the required viscous heating to set up an equilibrium
%state diverges ($\alpha\to\infty$). 
%In the present model the increased
%viscous damping, which is stabilizing, is balanced by the decreasing
%(and hence destablizing) $Q$ required to achieve large $\alpha$
%(Eq. \ref{Qalpha}).

{\bf maybe move following to discussion}
Numerical simulations of self-gravitating disks show there is a
maximum $\alpha$ ($\sim 0.06$) that can be sustained before
fragmentation \citep{rice11}. We can interpret this result in our
linear framework. 
%We can use our case (\ref{gvisc}) to interpret this
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$. Fig. \ref{gammie_maxrate_plot} 
shows that such a system would be dynamically unstable with 
growth rate $s = O(\Omega)$. This due to the direct effect of viscous 
stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 
and hence highly turbulent, self-gravitating disks to persist beyond
dynamical timescales. This is consistent with previous numerical
simulations performed by \cite{lodato05}. 
%Case (\ref{gvisc}) is consistent with 
\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    1.0cm]{figures/result2d_gvisc_kmax}
  \caption{Growth rates (top) at the optimal wavenumber (bottom) for
    viscous GI as a function of cooling time for the
    case shown in Fig. \ref{gammie_rate_plot}. Solid lines are
    computed from the dispersion relation (Eq. \ref{thindisk}), and the dotted line is an
    estimate from Eq. \ref{gammie_maxrate_simple}. The vertical dashed
    line marks the region with $\alpha > 1$, and the vertical
    dashed-dot line marks the region with growth timescales longer
    than dynamical. 
    \label{gammie_maxrate_plot}}
\end{figure}
